Matroids from families of null hypotheses

### 2013-08-16 presentation/cycle.tex
included first example graphs, changed scaling of graphs to .5
added cycles, rearranged slides, added lattice

\begin{tikzpicture}[scale=1]
\node (H1) at (225bp,-125bp)[draw,circle,fill=white] {$\mu_1$};
\node (H2) at (425bp,-125bp)[draw,circle,fill=white] {$\mu_2$};
\node (H3) at (425bp,-325bp)[draw,circle,fill=white] {$\mu_3$};
\node (H4) at (225bp,-325bp)[draw,circle,fill=white] {$\mu_4$};
\draw [line width=1pt] (H1) to (325bp, -124bp) node[fill=blue!20] {$H_{1}$} to (H2);
\draw [line width=1pt] (H1) to (275bp, -175bp) node[fill=blue!20] {$H_{2}$} to (H3);
\draw [line width=1pt] (H2) to (425bp, -224bp) node[fill=blue!20] {$H_{4}$} to (H3);
\draw [line width=1pt] (H3) to (326bp, -324bp) node[fill=blue!20] {$H_{6}$} to (H4);
\draw [line width=1pt] (H4) to (225bp, -227bp) node[fill=blue!20] {$H_{3}$} to (H1);
\draw [line width=1pt] (H4) to (374bp, -176bp) node[fill=blue!20] {$H_{5}$} to (H2);
\end{tikzpicture}

>>>>

\begin{tikzpicture}[scale=.5]
\node (H1) at (225bp,-125bp)[draw,circle,fill=white] {$\mu_1$};
\node (H2) at (425bp,-125bp)[draw,circle,fill=white] {$\mu_2$};
\node (H3) at (425bp,-325bp)[draw,circle,fill=white] {$\mu_3$};
\node (H4) at (225bp,-325bp)[draw,circle,fill=white] {$\mu_4$};
\only<2,5,6,8>{\draw [line width=1pt] (H1) to (325bp, -124bp) node[fill=blue!20] {$H_{1}$} to (H2);}
\only<2,4,7,8>{\draw [line width=1pt] (H1) to (275bp, -175bp) node[fill=blue!20] {$H_{2}$} to (H3);}
\only<2,3,6,7,9>{\draw [line width=1pt] (H2) to (425bp, -224bp) node[fill=blue!20] {$H_{4}$} to (H3);}
\only<3,4,6,8,9,10>{\draw [line width=1pt] (H3) to (326bp, -324bp) node[fill=blue!20] {$H_{6}$} to (H4);}
\only<4,5,6,7>{\draw [line width=1pt] (H4) to (225bp, -227bp) node[fill=blue!20] {$H_{3}$} to (H1);}
\only<3,5,7,8,9,10>{\draw [line width=1pt] (H4) to (374bp, -176bp) node[fill=blue!20] {$H_{5}$} to (H2);}
\only<9,10>{\draw [line width=2pt,color=red,dashed] (H1) to (325bp, -124bp) node[fill=blue!20] {$H_{1}$} to (H2);}
\only<9,10>{\draw [line width=2pt,color=red,dashed] (H1) to (275bp, -175bp) node[fill=blue!20] {$H_{2}$} to (H3);}
\only<9>{\draw [line width=2pt,color=red,dashed] (H2) to (425bp, -224bp) node[fill=blue!20] {$H_{4}$} to (H3);}

\end{tikzpicture}

### 2013-08-16 presentation/presentation.tex
added cycles, rearranged slides, added lattice

\begin{document}

@@

\begin{frame}
\frametitle{Example 1: 4 Means}
\input{full4}
\end{frame}

@@

\begin{itemize}
\item $H_i$ depends on $H_I$ iff $H_{I \cup i} = H_{I}$, induces a
  dependence relation $\sim_H$ on $\mathcal{P}(S)$
\item $H_I$ independent iff for all $i \in I$, $i \nsim_H I$,
\item $I$ independent $\Leftrightarrow$ $H_{I} \subsetneq  H_{I\setminus i}$
\item For pairwise comparisons $I$ independent iff there are no cycles in $\mathscr{G}_I$
\item For general contrast tests $I$ independent iff $\bs{C_I}$ has full
  column rank
\end{itemize} % ends low level
\end{frame}

@@

\item A matroid on $S$ endows $\mathcal{P}(S)$ with an algebraic
  structure 
\item Matroids generalize the structures of graphs and vector spaces,
  \eg circuits, bases, independent sets
\item Many ways to define a matroid
  \begin{itemize}
  \item From a graph perspective: Circuit axioms
  \item From a vector space perspective: Independence axioms
  \end{itemize}
\item There is a dual matroid $M^*$ which can be helpful for proofs and algorithms 
\item \textbf{/Some illustration hereabouts/}
\end{itemize} % ends low level
\end{frame}

@@

\begin{tabular}{ll}
  \begin{minipage}{.4\textwidth}
    \input{cycle}
  \end{minipage} &
  \begin{minipage}{.6\textwidth}
    Consider $S$ and a collection of subsets $\mathscr{C}$ for which holds
    \begin{enumerate}
    \item $C_1 \neq C_2 \in \mathscr{C}$, then $C_1 \nsubseteq C_2$,
    \item $C_1 \neq C_2 \in \mathscr{C}$, and $i \in C_1 \cap C_2$ then
      $\exists C_3 \subseteq (C_1 \cup C_2)$,
    \end{enumerate}
    then $\mathscr{C}$ is the set of circuits of a matroid $M$ on $S$
  \end{minipage}
\end{tabular}

@@

\begin{frame}{Maximally independent sets}
  \begin{itemize}
  \item Bases are maximally independent sets of $S$
  \item Rank $\rho I$ is the cardinality of a maximally indepedendent set in $I$
  \item Circuits are minimally dependent sets, \ie each subset of $C$
    with rank $\rho C - 1$ is a maximally independent set of $C$
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Maximally dependent sets}
\begin{itemize}
\item The closure of a set $\sigma I = \{i \in S: i \sim I\}$
\item Closed sets {\em flats}: $\mathscr{L} = \{I \subseteq S: \sigma I = I\}$
\item Hyperplanes are flats of rank $\rho S - 1$
\item All flats are given by closing hyperplanes under intersection
\item Circuits in $M^*$ are the complements of hyperplanes in $M$
\end{itemize} % ends low level
\end{frame}

@@

\section{Matroids from logically dependent hypotheses}

  \begin{block}{Theorem}
    For a family of two-sided pairwise mean comparisons $H_i$, $i \in
    S$, an intersection hypotheses $H_I$ is independent iff the
    corresponding edge set $\{e_i:i \in I\}$ is independent in the
    cycle matroid on $\mathscr{G}_I$.

For a family of two-sided
    general contrast tests $H_i$, $i \in S$, an intersection
    hypotheses $H_I$ is independent iff the corresponding contrast
    matrix $\bs{C}_I$ is independent in the vectorial matroid on
    $\mathbb{R}^m$.

    \begin{displaymath}
      i \sim_H I \Leftrightarrow i \sim_M I
    \end{displaymath}
  \end{block}

@@

\begin{frame}
\frametitle{Closed test procedure}
\begin{itemize}

\item Every elementary hypothesis is a flat

\item Using local level-$\alpha$ tests for each flat we have a closed test procedure that controls the FWER in the strong sense (Closed test principle)

\begin{frame}
\frametitle{Examples}
\begin{itemize}

\item Pairwise comparisons

\item The flats of the complete graph correspond to the partitions of
  $\{1,...,m\}$
\item This provides efficient methods to generate all flats for
  pairwise mean comparisons \cite{er1988fast,knuth2005art,kokosinski2006new}

\item The flats of a matroid generate a lattice, procedures testing not all possible comparisons correspond to intervals on that lattice

@@

\textbf{Old wine in new bottles ((c) L. Hothorn)}
\begin{itemize}
\item \cite{westfall1997multiple} checks each intersection hypotheses
  whether it is flat using dependence arguments.
\item \cite{westfall2007multiple} provide a branch and bound algorithm
  to generate all flats. They note that for pairwise comparisons all
  exhaustive hypotheses are {\em partition hypotheses} 
\item \cite{bergmann1988improvements} define the concepts of an
  exhaustive sets
\item \cite{bernhard1991computergestuetzte} shows properties of
  exhaustive sets that are basically {\em closure axioms} of a matroid
  and provides algorithms to check each intersection hypotheses if
  flat using rank function.

  \cite{hommel1999bonferroni} show {\em closure axioms} for exhaustive
  sets 
\item Flats <-> Weakly Exhaustive set (Hommel, Bergmann)

\item Lemma 1.5 in (Bernhard) about weakly exhaustive sets are ``closure axioms'' of a matroid

\item Graphical procedures for pairwise comparisons (Weichert) identify dependence using cycle in the graph

\item \emph{However:}
\begin{itemize}
\item proofs are shorter (more elegant),
\item efficient algorithms exist to find the flats of matroids,
\item we may define procedures using independent sets,
\item \ldots{}
\end{itemize}

\begin{frame}{Matroid procedures for pairwise mean comparisons}

\begin{tabular}{ll}
  \begin{minipage}{.3\textwidth}
    \includegraphics[width=1.1\textwidth]{setpartitions.pdf} \\
     {\tiny wikipedia:Bell\_number}
  \end{minipage} &
  \begin{minipage}{.7\textwidth}
  \begin{itemize}
  \item Two-sided comparisons between $m$ means
  \item The cycle matroid of the corresponding graph representation
    provides an equivalent dependence structure
  \item The flats of a (simple) complete graph are
    isomporphic to the partitions of a set of size $m$
  \item Provides an efficient way to generate all exhaustive sets by
    enumeration of all partitions or hyperplanes \cite{er1988fast,knuth2005art,kokosinski2006new} 
  \item The number of flats is given by the Bell-numbers \ie 5, 15,
    52, 203, 877, 4140, ... 
  \end{itemize}
  \end{minipage}
\end{tabular}

@@

  \end{minipage} &
  \begin{minipage}{.6\textwidth}
    \begin{itemize}
    \item Provides weights $w_{i,I}$ for each elementary hypotheses
      $H_i$ and each non-empty subset $I \subset S$
    \item Weights are used to define a closed test procedure of
      weighted intersection hypotheses tests
    \item We have $\sum_{i \in S} w_{i,I} \leq 1$ (conservative),
    \item and $w_{i,I} \leq w_{i,J}$ for $J \subseteq I$ (consonance).
    \end{itemize}
  \end{minipage}
\end{tabular}
\end{frame}

>>>>

\begin{document}
\frametitle{Matroids}

@@

% \begin{frame}
% \frametitle{Example 1: 4 Means}
% \input{full4}
% \end{frame}

@@

\begin{itemize}
\item $H_i$ depends on $H_I$ iff $H_{I \cup i} = H_{I}$, induces a
  dependence relation $\sim_H$ on $\mathcal{P}(S)$
\item $H_I$ independent iff for all $i \in I$, $i \nsim_H I$,
\item $I$ independent $\Leftrightarrow$ $H_{I} \subsetneq
  H_{I\setminus i}$ for all $i \in I$
\pause
\item For pairwise comparisons $I$ independent iff there are no cycles in $\mathscr{G}_I$
\item For general contrast tests $I$ independent iff $\bs{C_I}$ has full
  column rank
\end{itemize} % ends low level
\end{frame}

@@

\item Matroids generalize the structures of graphs and vector spaces,
  \eg cycles, bases, independent sets
\item A matroid $M$ on $S$ endows $\mathcal{P}(S)$ with an algebraic
  structure 
\item Cryptomorphism: many ways to define a matroid: 
  \begin{itemize}
  \item From a graph perspective: Circuit axioms.
  \item From a vector space perspective: Independence axioms.
  \end{itemize}
\item There is a dual matroid $M^*$ which can be helpful for proofs and algorithms 
\end{itemize} % ends low level
\end{frame}

@@

\begin{tabular}{ll}
  \begin{minipage}{.4\textwidth}
    \only<2->{\input{cycle}}
.
  \end{minipage} &
  \begin{minipage}{.6\textwidth}
    Consider $S$ and a collection of subsets $\mathscr{C}$ for which holds
    \begin{enumerate}
    \item $C_1 \neq C_2 \in \mathscr{C}$, then $C_1 \nsubseteq C_2$,
    \item $C_1 \neq C_2 \in \mathscr{C}$, and $i \in C_1 \cap C_2$ then
      $\exists C_3 \subseteq (C_1 \cup C_2)$,
    \end{enumerate}
    then $\mathscr{C}$ is the set of circuits of a matroid $M$ on $S$
  \end{minipage}
\end{tabular}

@@

\section{Matroids from logically dependent hypotheses}
\begin{frame}
  \frametitle{Dependent hypotheses}
  \begin{block}{Equivalence of dependence relations $\sim_H = \sim_M$}
    \begin{itemize}
    \item<1-> $H_i$ depends on $H_I$ iff $H_{I \cup i} = H_{I}$
    \item<2-> Does this mean $\mathscr{H}$ defines a matroid $M$, such that
      $\sim_H = \sim_M$?
    \end{itemize}
  \end{block}
\end{frame}

  \begin{block}{Two-sided pariwise comparisons}
    For a family of two-sided pairwise mean comparisons $H_i$, $i \in
    S$, an elementary hypotheses $H_j$ depends on intersection
    hypotheses $H_I$ iff the corresponding edge $e_j$ is dependent on
    the edge set $\{e_i:i \in I\}$ of the cycle matroid defined by 
    $\mathscr{G}_I$.   
  \end{block}

  \begin{block}{Two-sided linear contrast tests}
    For a family of two-sided general contrast tests $H_i$, $i \in S$,
    an elementary hypothesis $H_j$ depends on intersection hypotheses
    $H_I$ iff the corresponding contrast vector $\bs{c_j}$ is
    dependent on the column vectors of the matrix $\bs{C}_I$ in the
    vectorial matroid defined by $\bs{C}$ on $\mathbb{R}^m$.
  \end{block}

\begin{frame}
  \frametitle{Matroids from null hypotheses}
  {\Large What is it good for?\\
  \invisible<1>{Structure}}
\end{frame}

\begin{frame}
\frametitle{Maximally independent sets}
  \begin{itemize}
  \item Bases are maximally independent sets of $S$
  \item Rank $\rho I$ is the cardinality of a maximally indepedendent set in $I$
  \item Circuits are minimally dependent sets, \ie each subset of $C$
    with rank $\rho C - 1$ is a maximally independent set of $C$
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Maximally dependent sets}
\begin{itemize}
\item The closure of a set $\sigma I = \{i \in S: i \sim I\}$
\item Closed sets {\em flats}: $\mathscr{L} = \{\sigma I: I \subseteq S\}$
\item Hyperplanes are flats of rank $\rho S - 1$
\item All flats are given by closing hyperplanes under intersection
\item Circuits in $M^*$ are the complements of hyperplanes in $M$
\end{itemize} % ends low level
\end{frame}

@@

\begin{frame}
\frametitle{Flats}
\begin{itemize}

\item Every elementary hypothesis is a flat
\item For each $I \subseteq S$, $H_I = H_{\sigma I}$
\item Define a stepwise procedure on the flats of $M$

\begin{frame}
  \frametitle{Stepwise test procedure}
  \begin{block}{Multiple testing procedure for logically related hypotheses}
    Assume that $\mathscr{H}$ with dependence relation $\sim_H$
    defines a matroid on $M$ on $S$. Then a test procedure that
    rejects elementary hypotheses $H_i$ if all $H_E$, $i \in E$, $E
    \in \mathscr{L}$ are rejected using a local level-$\alpha$ test
    controls the FWER in the strong sense.
  \end{block}

@@

\begin{frame}
  \frametitle{Stepwise multiple testing procedure on the lattice of
    flats}

\begin{tabular}{ll}
  \begin{minipage}{.3\textwidth}
    \includegraphics<1>[width=1.2\textwidth]{lattice4.pdf}
    \includegraphics<2>[width=1.2\textwidth]{lattice4accept.pdf}
    \includegraphics<3>[width=1.2\textwidth]{lattice4reject.pdf}
    \includegraphics<4>[width=1.2\textwidth]{lattice4rejacc.pdf}\\
    {\tiny wikipedia:Partition\_of\_a\_Set}
  \end{minipage} &
  \begin{minipage}{.7\textwidth}
  \begin{itemize}
  \item All flats are given by intersections of hyperpanes
    $\rightarrow$ knowing all hyperplanes remaining flats are given by
    intersection of index sets
  \item Hyperplanes can by found by algorithms that enumerate the
    circuits of the dual matroid \cite{boros2003algorithms}
  \item The partial order of flats defined by inclusion defines a
    geometric lattice
  \item<2-> Testing only a subset of elementary hypotheses corresponds
    to intervals
  \item<3-> Rejection of elementary hypotheses removes intervals
  \end{itemize}

\begin{frame}{Matroid procedures for pairwise mean comparisons}
\begin{tabular}{ll}
  \begin{minipage}{.3\textwidth}
    \includegraphics[width=1.1\textwidth]{setpartitions.pdf} \\
     {\tiny wikipedia:Bell\_number}
  \end{minipage} &
  \begin{minipage}{.7\textwidth}
  \begin{itemize}
  \item Two-sided comparisons of $m$ means
  \item The cycle matroid of the graph representation
    provides equivalent dependence structure
  \item The flats of a (simple) complete graph are
    isomporphic to the partitions of a set of size $m$
  \item Efficient algorithms to generate all exhaustive sets by
    enumeration of set partitions \cite{er1988fast,knuth2005art,kokosinski2006new} 
  \item The number of flats are $m$-th Bellnumber $-1$ \ie 4, 14,
    51, 202, 876, 4139, ... 
  \end{itemize}
  \end{minipage}
\end{tabular}

@@

\textbf{Old wine in new bottles ((c) L. Hothorn)}
\begin{itemize}
\item \cite{shaffer1986modified} represents pairwise comparisons
  by set partitions 
\item \cite{bergmann1988improvements} define the concepts of an
  exhaustive sets
\item \cite{bernhard1991computergestuetzte} shows properties of
  exhaustive sets that are basically {\em closure axioms} of a matroid
  and provides algorithms to check each intersection hypotheses if
  flat using rank function.
\item \cite{westfall1997multiple} checks each intersection hypotheses
  whether it is flat using dependence arguments.
\item \cite{weichert2000robuste} suggest an MTP based on clique
  decomposition of the graph
\item \cite{westfall2007multiple} provide a branch and bound algorithm
  to generate all flats. They note that for pairwise comparisons all
  exhaustive hypotheses are {\em partition hypotheses} 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Matroid representation}
  \begin{itemize}
  \item proofs are shorter (more elegant),
  \item efficient algorithms exist to find the flats of matroids,
  \item {\em Scumm principle}\cite{welsh2010matroid} everything
    important happens at the top (hyperplanes)
  \item we may define procedures using independent sets,
  \item \ldots{}
  \end{itemize}
\end{frame}

@@

  \end{minipage} &
  \begin{minipage}{.6\textwidth}
    \begin{itemize}
    \item Provides weights $w_{i,I}$ for each index $i$ and each
      non-empty subset $I \subset S$
    \item Weights are used to define a closed test procedure of
      weighted intersection hypotheses tests
    \item We have $\sum_{i \in S} w_{i,I} \leq 1$ (conservative),
    \item and $w_{i,I} \leq w_{i,J}$ for $J \subseteq I$ (consonance).
    \end{itemize}
  \end{minipage}
\end{tabular}
\end{frame}
